---
title: Estimating extinction time from the fossil record accounting for measurement error
author: "David Warton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,echo = TRUE)
```

## Problem statement

Often we wish to estimate the time of extirpation of a species using the fossil record -- we have a sequence of estimated dates for the species, measured with error (whose magnitude is assumed to be known). For example, say we have the following data (from the Irish Elk, sourced from supplementary material for Cooper et al 2015; Science):
```{r data input}
deer = c(14215.25, 14234.5, 14160, 14095.5, 14059.75, 13984.5, 13884.75, 13864, 13854, 13844.5, 13675.5, 13664.25, 13624.25, 13643.5, 13589.75, 13576.5, 13490.75, 13445, 13431.25, 13399.5, 13379.25, 13382.75, 13357.5, 13348.5, 13347.25, 13340.25, 13258.75, 13192.75, 13192.75, 13189.5, 13182.25, 13033.5, 13022.25, 12956.5, 12825.75, 12839.75, 12860.25, 12791.25, 12779.75, 12693.25, 12628.75, 12739.75, 12591.5, 12432.75)
deerSds = c(105.75, 78.5, 69.5, 102, 131.25, 120, 97.25, 89.5, 89.5, 95, 78.5, 74.75, 132.75, 94, 116.75, 119, 54.75, 119.5, 112.75, 60, 90.75, 58.25, 93, 92, 65.75, 102.25, 284.25, 102.25, 62.75, 75, 95.75, 70, 55.75, 114, 74.75, 110.25, 205.75, 90.75, 97.25, 66.75, 78.25, 307.75, 68, 96.25)
```

There are two main sources of uncertainty to account for:

1. _Sampling error_, that is, the last time the species was seen (`r format(min(deer), scientific=FALSE)` years ago) is not necessarily the last time it was there.

1. _Measurement error_, because fossil ages were measured with uncertainty (whose standard deviation is assumed to be known in the below).

Currently, methods in the literature account for one or other of these quantities, but typically not both. Classical estimators (Strauss, McInerney's, _etc_) account for sampling error only, GRIWM constructs confidence intervals which account for measurement error but don't actually deal with sampling error in CI construction.

The `est_cutt` function in the `reginv` package accounts for both, and produces nearly unbiased estimates of extinction time and confidence intervals that are close to exact, when assumptions are satisfied.

## Using the `est_cutt` function

To use the `est_cutt` function from the `reginv` package, you need to specify as input:

- `ages`, a vector of estimated fossil ages
- `sd`, a vector of measurement error standard deviations (scalar would also be OK, assuming this is constant)
- `K`, the oldest time for a fossil to be considered for inclusion in this dataset. If it is not clear what this is you could always remove the oldest fossil and set `K` equal to its value.

There are other arguments you can play with as well, consult the help file for details.

```{r fit model}
# if required, devtools::install_github("eco-stats/reginv")
library(reginv)
deerFt = est_cutt(deer,deerSds,K=14250) # takes maybe a minute to run
deerFt
```

This returns a point estimate for extinction of `r format(deerFt$theta[2], scientific=FALSE)` BP, and by default, it also includes a 95% confidence interval for it, which in this case extends a couple of centuries either side of `r format(deerFt$theta[2], scientific=FALSE)` BP.

This code takes a fair while to run (maybe a minute?) because it is doing a lot of work in the background. By default, unless measurement error is large, it uses a new method called _regression inversion_ (`method="reginv"`) which computes a maximum likelihood estimator under the CUTT model (see below), then repeatedly simulates new data at different potential extinction times in order to find a range that is plausible considering the estimators obtained from simulated data.

A faster alternative method is to use inversion from the chi-squared distribution
```{r fit modelMLE}
est_cutt(deer,deerSds,K=14250,method="mle")
```

This is much faster because it doesn't simulate new values for the estimator, it instead assumes they are normally distributed (which is a reasonable assumption when neither sample size nor measurement error is small). This ended up returning similar values, but in cases where measurement error is small it tends to return confidence intervals that are too short.


## Model - the CUTT distribution

The `est_cutt` function uses a new technique based on modelling fossil ages using a Compound Uniform -- Truncated T (CUTT) distribution. Using this distribution assumes

- That, for a fossil of a given age, measurement error estimating its age follows a distribution that is Student's T multiplied by the provided standard deviation, truncated such that observed age is less than K. Unless it is specified, the degrees of freedom of the T distribution are estimated from the data (and need not be a whole number)

- That deposition rates of fossils is uniformly distributed over the interval of allowable dates

A truncated normal distribution could also be used (and is a special case of the T, where degrees of freedom is infinite) but we have noticed that measurement error seems to be long-tailed relative to the normal so the T tends to provide a better fit to data. Departures from uniform deposition rate could also be considered but usually I think there is not enough information to estimate a plausible alternative.

Considering our Irish Elk dataset:
```{r hist}
hist(deer)
```

While this looks fairly symmetrical, the model we are fitting actually expects this distribution to be left-skewed and shaped more like the below:

```{r dcutt}
xs=seq(12300,14250,length=100)
ds = dcutt(xs,12500,K=14250,sd=mean(deerSds),df=Inf)
plot(xs,ds,type="l")
```

The left-tail comes from measurement error. The assumed extinction time for this particular distribution is about halfway down this curve at 12500. The curve flattens out to the right away from the extinction time, due to the uniform assumption on deposition rates.

It can be hard to compare a sample histogram to a density curve to check assumptions, unless you have a very big sample size. I would suggest constructing a qqplot against the best-fitting CUTT distribution as follows:

```{r qqplot,fig.width=6}
qqenvelope(deerFt)
```

The shaded region is a 95% global envelope, and if data followed the CUTT distribution with parameters as stored in `deerFt` then we would expect all points to be inside it. In the above plot we have all points inside, and generally close to the one-to-one line, suggesting no evidence against the CUTT distribution in this case.
