---
title: Estimating extinction time from the fossil record accounting for measurement error
author: "David Warton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,echo = TRUE)
```

## Problem statement

Often we wish to estimate the time of extirpation of a species using the fossil record -- we have a sequence of estimated dates for the species, measured with error (whose magnitude is assumed to be known). For example, say we have the following data (from a red deer sub-species, sourced from supplementary material for Cooper et al 2015; Science):
```{r data input}
x=c(14729, 14903, 14743, 14738, 14739, 14741, 14732, 14720, 14657, 14638, 14635, 14631, 14642, 14613, 14577, 14544, 14539, 14545, 14541, 14541, 14507, 14490, 14496, 14440, 14389, 14389, 14413, 14361, 14391, 14373, 14323, 14278, 14306, 14301, 14229, 14101, 14068, 14035, 14057, 14018, 13999, 13951, 13939, 13896, 13835, 13819, 13660, 13485, 13482, 12714, 12514, 12227, 12202, 12122, 12124)
sds=c(333, 176, 305, 301, 282, 278, 264, 261, 286, 309, 311, 295, 245, 256, 293, 252, 259, 302, 290, 289, 282, 240, 299, 279, 224, 230, 287, 216, 273, 284, 299, 251, 293, 301, 392, 268, 253, 244, 128, 161, 203, 175, 115, 158, 194, 207, 85, 98, 78, 55, 115, 179, 186, 243, 252)
```

There are two main sources of uncertainty to account for:

1. _Sampling error_, that is, the last time the species was seen (12122 years ago) is not necessarily the last time it was there.

1. _Measurement error_, because fossil ages were measured with uncertainty (whose standard deviation is assumed to be known in the below).

Currently, methods in the literature account for one or other of these quantities, but typically not both. Classical estimators (Strauss, McInerney's, _etc_) account for sampling error only, GRIWM constructs confidence intervals which account for measurement error but not sampling error.

The `est_cutt` function in the `reginv` package accounts for both, and produces nearly unbiased estimates of extinction time and confidence intervals that are close to exact.

## Using the `est_cutt` function

To use the `est_cutt` function from the `reginv` package, you need to specify as input:

- `ages`, a vector of estimated fossil ages
- `sd`, a vector of measurement error standard deviations (scalar would also be OK, assuming this is constant)
- `K`, the oldest time for a fossil to be considered for inclusion in this dataset. If it is not clear what this is you could always remove the oldest fossil and set `K` equal to its value.

There are other arguments you can pay with as well, consult the help file for details.

```{r fit model}
# if required, devtools::install_github("dwarton/reginv")
library(reginv)
est_cutt(x,sds,K=15000) # takes maybe a minute to run
```

This returns a point estimate for extinction of 12000 or so, and by default, it also includes a 95% confidence interval for it, which in this case extends a few centuries either side of 12000 BP.

This code takes a fair while to run (maybe a minute?) because it is doing a lot of work in the background. By default, unless measurement error is large, it uses a new method I developed called _regression inversion_ (`method="reginv"`) which computes a maximum likelihood estimator under the CUTT model (see below), then repeatedly simulates new data at different potential extinction times in order to find a range that is plausible considering the estimators obtained from simulated data.

A faster alternative method is to use inversion form the chi-squared distribution
```{r fit modelMLE}
# if required, devtools::install_github("dwarton/reginv")
library(reginv)
est_cutt(x,sds,K=15000,method="mle")
```

This ended up returning similar values, but in cases where measurement error is small it tends to return confidence intervals that are too short.


## Model - the CUTT distribution

The `est_cutt` function uses a new technique based on modelling fossil ages using what I'm calling a Compound Uniform -- Truncated T (CUTT) distribution. Using this distribution assumes

- That, for a fossil of a given age, measurement error estimating its age follows a distribution that is Student's T multiplied by the provided standard deviation, truncated such that observed age is less than K

- That fossil dates are uniformly distributed over the interval of allowable dates

A truncated normal distribution could also be used (and is a special case of the T, where degrees of freedom is infinite) but I have noticed that measurement error seems to be long-tailed relative to the normal so the T tends to provide a better fit to data. Departures from uniform deposition rate could also be considered but usually I think there is not enough information to estimate a plausible alternative.

```{r dat}
x=c(29705,
29907,
30725,
30726,
30794,
30803,
30823,
30996,
31062,
31699,
32865,
33855,
35976,
36299,
36318,
36462,
36819,
37129,
37928,
38326,
39773,
39683,
40450,
40914,
41006,
41043,
41402,
41468,
41771,
41967,
42141,
42016,
42223,
42236,
42565,
42377,
42701,
42717,
42736,
42817,
42846,
43195,
43204,
43359,
44069,
44133,
44192,
43785,
44328,
44363,
44465,
44938,
44571,
45020,
45333,
45417,
45944,
46083,
46425,
46299,
48278,
48281,
48298,
48599,
48501,
49285,
49417,
49449,
49443,
48771
)
sds=c(
  173,
169,
993,
998,
919,
900,
893,
847,
763,
1087,
2323,
711,
674,
1005,
330,
429,
1519,
1052,
980,
666,
1431,
1851,
670,
468,
387,
497,
707,
656,
597,
414,
690,
464,
622,
648,
296,
801,
468,
415,
407,
505,
548,
736,
761,
756,
1205,
876,
836,
406,
804,
908,
997,
1563,
1106,
1539,
1167,
1194,
783,
763,
2290,
994,
2917,
2886,
2868,
2741,
2749,
2042,
1884,
1919,
2095,
2916
)
```

For example:
```{r hist}
hist(x,breaks=10)
```

Note that the data start off looking fairly flat then develop a strong left-tail. This is what we would expect from a CUTT distribution:

```{r dcutt}
# if required, devtools::install_github("dwarton/reginv")
library(reginv)
xs=seq(25000,50000,length=100)
ds = dcutt(xs,29500,K=50000,sd=mean(sds),df=3)
plot(xs,ds,type="l")
```

The flat bit to the right is the uniform part, and the left-tail comes from measurement error. The assumed extinction time for this particular distribution is halfway down this curve at 13000.

Comparing this curve with the actual data, I think the left-tail for the observed data is longer than we would expect due to measurement error -- this means either that deposition rate was not constant, but tailed off towards extinction time, or it means that measurement error is larger than the values suggested in input.

```{r qqplot}
# install.packages("ecostats")
ecostats::qqenvelope(qnorm(pcutt(x,ft$theta[2],sd=sds,K=50000,df=ft$df)))
```